The Impact of the 2022-23 NBA Rule Change

### 2024-04-23 paper/paper.qmd
Update model and results section
Add data analysis

- $Y$ is the MVP Voting Points. It is the dependent variable that represents the points a player receives in the MVP voting process.
- $\beta_0$ is the intercept of the regression model that represents the expected value of Total MVP Voting Points when all predictor variables (FG%, 3P%, FT%) are at zero.
- $\beta_1$ is the coefficient for FG%. It indicates how changes in a player’s overall shooting efficiency from the field affect their MVP voting points.
- $\beta_2$ is the the coefficient for 3P%. This reflects the impact of three-point shooting efficiency on MVP voting.
- $\beta_3$ is the coefficient for FT%. This measures the influence of free throw shooting efficiency on MVP voting points.
- $X_1$ measures the FG% made out of attempts.
- $X_2$ measures the 3P% made out of attempts.
- $X_3$ measures the FT% made out of attempts.
- $\epsilon$ is the error term, representing the residual element not explained by the model. This captures the variability in MVP voting points not explained by shooting percentages alone.

@@

# Results {#sec-results}

Our results are summarized in @tbl-coefficients and @tbl-additional (see @fig-modelresults)

@@

# Generate a tidy summary of the model
model_summary <- tidy(model2023) |>
  rename(
    Term = term,
    Estimate = estimate,
    `Std. Error` = std.error,
    Statistic = statistic,
    `P-value` = p.value,
  ) |>
  # Rename the "Term" values
  mutate(
    Term =
      case_match(
        Term,
        "(Intercept)" ~ "(Intercept)",
        "field_goal_percentage" ~ "Field Goal Percentage",
        "three_points_percentage" ~ "Three Points Percentage",
        "free_throw_percentage" ~ "Free Throw Percentage"
        )
    )

@@

@@

@@

@@

# Diagnostic plots
#par(mfrow = c(2, 2))  # Arrange plots in a 2x2 grid
#plot(model2023)  # Generate plots

@@

\appendix
# Appendix {-}

# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

>>>>

- $Y$ is the MVP Voting Points. It is the dependent variable that represents the points a player receives in the MVP voting process.
- $X_1$ measures the FG% made out of attempts.
- $X_2$ measures the 3P% made out of attempts.
- $X_3$ measures the FT% made out of attempts.
- $\beta_0$ is the intercept of the regression model that represents the expected value of Total MVP Voting Points when all predictor variables (FG%, 3P%, FT%) are at zero.
- $\beta_1$ is the coefficient for FG%. It indicates how changes in a player’s overall shooting efficiency from the field affect their MVP voting points.
- $\beta_2$ is the the coefficient for 3P%. This reflects the impact of three-point shooting efficiency on MVP voting.
- $\beta_3$ is the coefficient for FT%. This measures the influence of free throw shooting efficiency on MVP voting points.
- $\epsilon$ is the error term, representing the residual element not explained by the model. This captures the variability in MVP voting points not explained by shooting percentages alone.

@@

# Results {#sec-results}
## Summary of regression model results
Our results are summarized in @tbl-coefficients and @tbl-additional. The estimate intercept is -920.99 with a high standard error of 1833.86 and high p-value of 0.63 (significantly greater than 0.05). The negative intercept might initially suggest a baseline negative effect, but considering high standard error and p-value, the intercept is not statistically significant. This means that the intercept does not provide reliable information about the baseline level of MVP points when shooting percentages are not accounted for.

Field Goal Percentage (FG%) has a positive coefficient, indicating that an increase in FG% will have an increase in MVP points. The estimate suggests that for each percentage point increase in FG%, MVP points increase by approximately 3432.88. However, the standard error of 2139, and the p-value of 0.160, which is above the significant level of 0.05, suggests that this relationship is not statistically significant at the conventional level. Thus, while FG% shows a positive impact on MVP points, this effect is not strong enough to be considered reliable within this model.

The coefficient for Three Points Percentage (3P%) is negative (-6037.23), suggesting that a higher 3P% correlates with fewer MVP points. It indicates a substantial decrease in MVP points for each percentage increase in 3P%. However, similar to FG%, the p-value of 0.140 indicates that this negative relationship is not statistically significant. The large standard error further points to high variability in the data, leading to uncertainty in the estimate.

For Free Throw Percentage (FT%), the regression model estimates a coefficient of 1828, which shows a positive relationship between FT% and MVP points. This suggests that higher FT% could be associated with more MVP points. However, the p-value of 0.356 indicates that this effect is not statistically significant, which exceeds the commonly accepted 0.05 for declaring statistical significance. Moreover, the standard error (1828.51) is nearly equal to the estimate (1827.86), indicating low precision in the estimate. Consequently, there is considerable uncertainty around its effect on MVP points.

@@

# Generate a tidy summary of the model
model_summary <- tidy(model2023) |>
  rename(
    Term = term,
    Estimate = estimate,
    `Std. Error` = std.error,
    Statistic = statistic,
    `p-value` = p.value,
  ) |>
  # Format numerical columns to the second decimal place
  mutate(across(where(is.numeric), ~format(round(., 2), nsmall = 2))) |>
  # Rename the "Term" values
  mutate(
    Term =
      case_match(
        Term,
        "(Intercept)" ~ "(Intercept)",
        "field_goal_percentage" ~ "Field Goal Percentage",
        "three_points_percentage" ~ "Three Points Percentage",
        "free_throw_percentage" ~ "Free Throw Percentage"
        )
    )

@@

@tbl-additional provides several important statistical metrics that help evaluate the overall fit and effectiveness of the regression model. The high Residual Standard Error of 314.30 suggests substantial unexplained variability, indicating the model's predictions can deviate significantly from actual values. While the Multiple R-squared value of 0.44 shows that the model explains about 44% of the variability in MVP points, the lower Adjusted R-squared of 0.17 indicates that the model's explanatory power significantly reduces when adjusted for the number of predictors, suggesting potential overfitting or irrelevance of some variables. Furthermore, the F-statistic of 1.60, which is relatively low, and p-value of 0.29 indicate that the model is not statistically significant, implying that the predictors do not reliably explain variations in MVP points.

@@

# Round the Value column to the second decimal place
additional_metrics$Value <- round(as.numeric(additional_metrics$Value), 2)

@@

## Observed versus predicted MVP points won
@fig-modelresults shows the observed MVP points (actual outcomes) against predicted MVP points (outcomes estimated by the model). The perfect predictions would result in all points falling exactly on the red dashed line, where the predicted values equal the observed values. However, as we can see, the data points are notably distant from this line, particularly at higher observed points. For instance, an observed value of 915 points in a predicted value of only 534.42. This significant deviation indicates that the model tends to underpredict MVP points for players with higher observed scores. In addition, for lower observed points, the model's performance is inconsistent. For instance, when the observed value is at 30, the predicted value is slightly negative (-5.09), which doesn't align with the possible range of MVP points. Conversely, an observed value of 3 has an inaccurate predicted value of 369.98. This means there is a potential issue of reliability and precision in the model’s predictions for top-performing players.

@@

@@
